---
title: "LMM analyses assumption check - TeenLight"
author: "R Lazar"
date: '`r format(Sys.time(), "%d %B, %Y")`'
format:
  html:
    embed-resources: false
    code-fold: true
    code-summary: "Show the code"
editor: source
warning: false
editor_options:
  chunk_output_type: console
---


# Preparing the environment

```{r, include=FALSE}

#Clear existing data and graphics
rm(list=ls())
graphics.off()

```

## Loading the Libraries

```{r Load libraries, warning=FALSE}

# Install and load pacman if not already installed
if (!require("pacman")) install.packages("pacman")

# Load all the packages using pacman
pacman::p_load(
  #for LMM
  lme4,
  lmerTest,
  sjPlot,
  
  # for general data wrangling
  DescTools,
  lubridate,
  tidyverse,
  
  qqplotr,    # For QQ plots in linear assumptions check
  performance, # For linear assumptions check
  see,        # For linear assumptions check
  rmarkdown,   # For reproducible reporting
  here# for loading data easily
)

```



# Melatonin analysis

## Evening melatonin AUC (primary outcome)

-    Here we use the evening AUC computed with the trapedzoid method.
-    n=20 (2 participants excluded)


```{r load data and functions, echo=FALSE}

load(file = here("Datasets", "Mel_summary.rda"))

#exclude 2 participants
participants_to_exclude_no <- c("16", "17")

Mel_summary <- Mel_summary %>%
  filter(!record_id %in% participants_to_exclude_no)

```

### Defining LMMs:

- covariate model (additionally including bright light history, pubertal stage and chronotype)


```{r }
AUC_mel_eve <- lmer(
formula =  auc_mel_eve ~ Block + D1_TAT1k  + mctq_sc4_chronotype + pub_stage + (1 |record_id),
                data = Mel_summary) 


```

- sparse model

```{r }
AUC_mel_eve_sparse <- lmer(
formula =  auc_mel_eve ~ Block + (1 | record_id),
                data = Mel_summary) 

```

### Assumption check covariate model

```{r,  fig.height = 10, fig.width = 15, fig.align = "center"}

assumptest_mel <- performance::check_model(AUC_mel_eve, check=c("pp_check", "ncv",  "qq", "homogeneity","normality"),
                                           theme="ggplot2::theme_classic", panel=T) #"outliers",  "normality",  "vif",
#base_size=6, title_size=6, axis_title_size=6, 

assumptest_mel

#save pdf in 6 x 8 inches in Landscape


#plot(check_normality(AUC_mel_eve), type="pp")

res <- resid(AUC_mel_eve)

res_df <- as.data.frame(res)

res_MELAUC <- ggplot(res_df, aes(x=res)) + 
  geom_histogram(bins=20, color="black", fill="grey")+
  labs(x="Residuals (Melatonin AUC)", y="Frequency")+
  theme_classic() #binwidth=6

res_MELAUC

ggsave(res_MELAUC, file ="C:/Users/Rafael Lazar/Documents/Git Projects/TeenLight/08_output/Melatonin/res_MELAUC.tiff", 
       width = 8, height = 6, units = "cm", dpi = 600)


check_collinearity(AUC_mel_eve)

check_normality(AUC_mel_eve)



```


## Melatonin Onset calculations

-    Here we use data  from fitted melatonin curves resulting from the "HockeyStick" fitting method.
-    n=20 (2 participants excluded)


```{r Hockey Stick analysis, echo=FALSE}
### Fitted melatonin curves with HockeyStick----------------------------------------------------

load(file = here("Datasets", "merged_mel_timing_HS.rda"))

#exclude 2 participants
merged_mel_timing_HS <- merged_mel_timing_HS %>% filter(!is.na(HS_time))


```

### Defining LMMs:

- covariate model (additionally including bright light history, pubertal stage and chronotype)


```{r }
HS_MLMO <- lmer(
formula =  HS_time ~ Block +  D1_TAT1k + mctq_sc4_chronotype + pub_stage + (1 | record_id),
                data = merged_mel_timing_HS) #  

```

- sparse model 

```{r }
HS_MLMO_sparse <- lmer(
formula =  HS_time ~ Block + (1 | record_id),
                data = merged_mel_timing_HS) #  

```


### Assumption check covariate model

```{r,  fig.height = 10, fig.width = 15, fig.align = "center"}

assumptest_MO <- check_model(HS_MLMO, check=c("pp_check", "ncv",  "qq", "homogeneity","normality"),                          theme="ggplot2::theme_classic") #"outliers",  "normality",  "vif",
#base_size=6, title_size=6, axis_title_size=6, 

assumptest_MO

res <- resid(HS_MLMO)

res_df <- as.data.frame(res)

res_MELAUC <- ggplot(res_df, aes(x=res)) + 
  geom_histogram(bins=20, color="black", fill="grey")+
  labs(x="Residuals (Melatonin Onset)", y="Frequency")+
  theme_classic() #binwidth=6

res_MELAUC
#save pdf in 6 x 8 inches in Landscape

check_collinearity(HS_MLMO)

check_normality(HS_MLMO)

#plot(check_normality(HS_MLMO), type="pp")


```

## Morning melatonin AUC (as indicator for mel offset)

 - This is included in the Supplementary materials not in the main results.
 - n=20 (2 participants excluded + 1 morning AUC = NA)


### Defining LMMs


```{r }

# Define  the models with different combinations of covariates
covariates_combinations_mor <- list(
  "model0" = auc_mel_mor ~ Block + (1 | record_id),
  "model7" = auc_mel_mor ~ Block + (1 | record_id) + mctq_sc4_chronotype + pub_stage + D1_TAT1k
)

```

```{r }
# Fit the models
model_results_mor <- lapply(covariates_combinations_mor, function(formula) {
  lmer(formula, data = Mel_summary)
})


```

### Assumption check covariate model

```{r,  fig.height = 10, fig.width = 15, fig.align = "center"}

assumptest_mor2 <- check_model(model_results_mor$model7, check=c("pp_check", "ncv",  "qq", "homogeneity","normality"),                          theme="ggplot2::theme_classic") #"outliers",  "normality",  "vif",
#base_size=6, title_size=6, axis_title_size=6, 

assumptest_mor2


res <- resid(model_results_mor$model7)

res_df <- as.data.frame(res)

res_MELAUC_mor_cov <- ggplot(res_df, aes(x=res)) + 
  geom_histogram(bins=20, color="black", fill="grey")+
  labs(x="Residuals (morning melatonin AUC)", y="Frequency")+
  theme_classic() #binwidth=6

res_MELAUC_mor_cov

```


# KSS analysis

- n=4 participants excluded from KSS analysis, as they only had a third of the recorded data.
- in the KSS analysis, we use factor "time" in the LMM models (also in interaction with the condition ("Block")


```{r KSS analysis, echo=FALSE}


load(file = here("Datasets", "eve_KSS.rda"))

load(file = here("Datasets", "arvo_KSS.rda"))

```



## Defining LMMs:

- define 4 models: 1 covariate and  1sparse model each for analyses during the afternoon light intervention and the later evening light condition. 

- covariate models additionally include bright light history, pubertal stage and chronotype.


```{r}

KSS_eve <- lmer(
formula =  kss ~ Block * centered_time + D1_TAT1k + mctq_sc4_chronotype  + pub_stage + (1 | record_id),
                data = eve_KSS) #  




KSS_arvo <- lmer(
formula =  kss ~ Block * centered_time + D1_TAT1k + mctq_sc4_chronotype  + pub_stage + (1 | record_id),  data = arvo_KSS) #  



KSS_eve_sparse <- lmer(
formula =  kss ~ Block * centered_time + (1 | record_id),
                data = eve_KSS) #  




KSS_arvo_sparse <- lmer(
formula =  kss ~ Block * centered_time + (1 | record_id),  data = arvo_KSS) #  



```



## KSS in the later evening condition


### Assumption check covariate model


```{r,  fig.height = 10, fig.width = 15, fig.align = "center"}

assumptest_KSS_eve <- check_model(KSS_eve, check=c("pp_check", "ncv",  "qq", "homogeneity","normality"),                          
                                  theme="ggplot2::theme_classic") #"outliers",  "normality",  "vif",
#base_size=6, title_size=6, axis_title_size=6, 

#plot(check_normality(KSS_eve), type="pp")

assumptest_KSS_eve


res <- resid(KSS_eve)

res_df <- as.data.frame(res)


ggplot(res_df, aes(x=res)) + 
  geom_histogram(bins=20, color="black", fill="grey")+
  labs(x="Residuals (evening KSS)", y="Frequency")+
  theme_classic() #binwidth=6


#save pdf in 6 x 8 inches in Landscape

check_collinearity(KSS_eve)

check_normality(KSS_eve)


```

## KSS during the afternoon to evening light intervention

### Assumption check covariate model


```{r,  fig.height = 10, fig.width = 15, fig.align = "center"}

assumptest_KSS_arvo <- check_model(KSS_arvo, check=c("pp_check", "ncv",  "qq", "homogeneity","normality"),                          
                                   theme="ggplot2::theme_classic") #"outliers",  "normality",  "vif",
#base_size=6, title_size=6, axis_title_size=6, 

#plot(check_normality(KSS_arvo), type="pp")
assumptest_KSS_arvo

res <- resid(KSS_arvo)


res_df <- as.data.frame(res)


ggplot(res_df, aes(x=res)) + 
  geom_histogram(bins=20, color="black", fill="grey")+
  labs(x="Residuals (afternoon KSS)", y="Frequency")+
  theme_classic() #binwidth=6





#save pdf in 6 x 8 inches in Landscape

check_collinearity(KSS_arvo)

check_normality(KSS_arvo)


```



# PVT analysis
-  PVT mean response speed (mean 1/RT) was the outcome of interest
- n=22 (no participants excluded)
- in the PVT analysis, we use factor "time" in the LMM models (also in interaction with the condition ("Block")


```{r, echo=FALSE}


load(file = here("Datasets", "eve_PVT_stat.rda"))

load(file = here("Datasets", "arvo_PVT_stat.rda"))


datastat_eve_mean_1divRT <- subset(eve_PVT_stat, variable == "mean_1divRT")

datastat_arvo_mean_1divRT <- subset(arvo_PVT_stat, variable == "mean_1divRT")


count_df <- datastat_eve_mean_1divRT %>%
  group_by(record_id,cond) %>%
  summarize(count = n())


```


## Defining LMMs:

- define 4 models: 1 covariate and  1sparse model each for analyses during the afternoon light intervention and the later evening light condition. 

- covariate models additionally include bright light history, pubertal stage and chronotype.


```{r}

PVT_mean_1divRT_eve <- lmer(
  formula =  value ~ cond * centered_time + D1_TAT1k + mctq_sc4_chronotype + pub_stage + 
    (1 | record_id), data = datastat_eve_mean_1divRT) #  




PVT_mean_1divRT_arvo <-  lmer(
  formula =  value ~ cond * centered_time + D1_TAT1k + mctq_sc4_chronotype + pub_stage + 
    (1 | record_id),  data = datastat_arvo_mean_1divRT) 



PVT_mean_1divRT_eve_sparse <- lmer(
  formula =  value ~ cond * centered_time  + (1 | record_id),
  data = datastat_eve_mean_1divRT)  


PVT_mean_1divRT_arvo_sparse  <-  lmer(
  formula =  value ~ cond * centered_time + (1 | record_id),  data = datastat_arvo_mean_1divRT) 


```

## PVT in the later evening condition


### Assumption check covariate model


```{r,  fig.height = 10, fig.width = 15, fig.align = "center"}

assumptest_PVT_eve <- check_model(PVT_mean_1divRT_eve,
                                  check=c("pp_check", "ncv",  "qq", "homogeneity","normality"), 
                                  theme="ggplot2::theme_classic") #"outliers",  "normality",  "vif",
#base_size=6, title_size=6, axis_title_size=6, 

#plot(check_normality(PVT_mean_1divRT_eve), type="pp")
assumptest_PVT_eve

res <- resid(PVT_mean_1divRT_eve)


res_df <- as.data.frame(res)


ggplot(res_df, aes(x=res)) + 
  geom_histogram(bins=20, color="black", fill="grey")+
  labs(x="Residuals (evening PVT response speed)", y="Frequency")+
  theme_classic() #binwidth=6


#save pdf in 6 x 8 inches in Landscape

check_collinearity(PVT_mean_1divRT_eve)

check_normality(PVT_mean_1divRT_eve)


```


## PVT during the afternoon to evening light intervention

### Assumption check covariate model


```{r,  fig.height = 10, fig.width = 15, fig.align = "center"}

assumptest_PVT_arvo<- check_model(PVT_mean_1divRT_arvo, 
                                  check=c("pp_check", "ncv",  "qq", "homogeneity","normality"),   
                                  theme="ggplot2::theme_classic") #"outliers",  "normality",  "vif",
#base_size=6, title_size=6, axis_title_size=6, 

#plot(check_normality(PVT_mean_1divRT_arvo), type="pp")

assumptest_PVT_arvo

res <- resid(PVT_mean_1divRT_arvo)


res_df <- as.data.frame(res)


ggplot(res_df, aes(x=res)) + 
  geom_histogram(bins=20, color="black", fill="grey")+
  labs(x="Residuals (afternoon PVT response speed)", y="Frequency")+
  theme_classic() #binwidth=6


#save pdf in 6 x 8 inches in Landscape

check_collinearity(PVT_mean_1divRT_arvo)

check_normality(PVT_mean_1divRT_arvo)


```


# DPG analysis

- DPG in 30 minute bins
- n=21 (1 participant excluded)
- in the DPG analysis, we use factor "time" in the LMM models (also in interaction with the condition ("Block")



```{r, echo=FALSE }


load(file = here("Datasets", "eve_DPG.rda"))

load(file = here("Datasets", "arvo_DPG.rda"))

```


## Defining LMMs:

- define 4 models: 1 covariate and  1sparse model each for analyses during the afternoon light intervention and the later evening light condition. 

- covariate models additionally include bright light history, pubertal stage and chronotype.


```{r }

DPG_eve <-  lmer(
formula =  DPG ~ Block * centered_time + D1_TAT1k + mctq_sc4_chronotype + pub_stage + (1|record_id),
                data = eve_DPG) # 


DPG_eve_sparse <-  lmer(
formula =  DPG ~ Block * centered_time + (1|record_id),
                data = eve_DPG) # 



DPG_arvo <-  lmer(
formula =  DPG ~ Block * centered_time + D1_TAT1k + mctq_sc4_chronotype  + pub_stage + (1|record_id),  data = arvo_DPG) #  



DPG_arvo_sparse <-  lmer(
formula =  DPG ~ Block * centered_time + (1|record_id),  data = arvo_DPG) #  


```

## DPG in the later evening condition


### Assumption check covariate model


```{r,  fig.height = 10, fig.width = 15, fig.align = "center"}

assumptest_DPG_eve<- check_model(DPG_eve, 
                                 check=c("pp_check", "ncv",  "qq", "homogeneity","normality"),               
                                 theme="ggplot2::theme_classic") #"outliers",  "normality",  "vif",
#base_size=6, title_size=6, axis_title_size=6, 

#plot(check_normality(DPG_eve), type="pp")

assumptest_DPG_eve

res <- resid(DPG_eve)

res_df <- as.data.frame(res)


ggplot(res_df, aes(x=res)) + 
  geom_histogram(bins=20, color="black", fill="grey")+
  labs(x="Residuals (evening DPG)", y="Frequency")+
  theme_classic() #binwidth=6


#save pdf in 6 x 8 inches in Landscape

check_collinearity(DPG_eve)

check_normality(DPG_eve)


```

## DPG during the afternoon to evening light intervention

### Assumption check covariate model


```{r,  fig.height = 10, fig.width = 15, fig.align = "center"}

assumptest_DPG_arvo<- check_model(DPG_arvo,
                                  check=c("pp_check", "ncv",  "qq", "homogeneity","normality"), 
                                  theme="ggplot2::theme_classic") #"outliers",  "normality",  "vif",
#base_size=6, title_size=6, axis_title_size=6, 

#plot(check_normality(DPG_arvo), type="pp")

assumptest_DPG_arvo

res <- resid(DPG_arvo)

res_df <- as.data.frame(res)


ggplot(res_df, aes(x=res)) + 
  geom_histogram(bins=20, color="black", fill="grey")+
  labs(x="Residuals (afternoon DPG)", y="Frequency")+
  theme_classic() #binwidth=6


#save pdf in 6 x 8 inches in Landscape

check_collinearity(DPG_arvo)

check_normality(DPG_arvo)


```



