---
title: "LMM analyses TeenLight"
author: "R Lazar"
date: '`r format(Sys.time(), "%d %B, %Y")`'
format:
  html:
    embed-resources: true
    code-fold: false
    code-summary: "Show the code"
editor: source
warning: false
editor_options:
  chunk_output_type: console
---


# Preparing the environment

```{r, include=FALSE}

#Clear existing data and graphics
rm(list=ls())
graphics.off()

```

## Loading the Libraries

```{r Load libraries, warning=FALSE}

# Install and load pacman if not already installed
if (!require("pacman")) install.packages("pacman")

# Load all the packages using pacman
pacman::p_load(
  #for LMM
  lme4,
  lmerTest,
  sjPlot,
  
  # for general data wrangling
  DescTools,
  lubridate,
  tidyverse,
  
  qqplotr,    # For QQ plots in linear assumptions check
  performance, # For linear assumptions check
  see,        # For linear assumptions check
  rmarkdown,   # For reproducible reporting
  here# for loading data easily
)

```

---

# Melatonin analysis

## Evening melatonin AUC (primary outcome)

-    Here we use the evening AUC computed with the trapedzoid method.
-    n=20 (2 participants excluded)


```{r load data and functions, echo=FALSE}

load(file = here("LMM analysis/Datasets", "Mel_summary.rda"))

#exclude 2 participants
participants_to_exclude_no <- c("16", "17")

Mel_summary <- Mel_summary %>%
  filter(!record_id %in% participants_to_exclude_no)

```


### Defining LMMs:

- covariate model (additionally including bright light history, pubertal stage and chronotype)


```{r }
AUC_mel_eve <- lmer(
formula =  auc_mel_eve ~ Block + D1_TAT1k  + mctq_sc4_chronotype + pub_stage + (1 |record_id),
                data = Mel_summary) 


```

- sparse model

```{r }
AUC_mel_eve_sparse <- lmer(
formula =  auc_mel_eve ~ Block + (1 | record_id),
                data = Mel_summary) 

```

### Compute LMM results for covariate model (relative to "dim" condition)

```{r }


summary(AUC_mel_eve)

confint(AUC_mel_eve)

```

### Random effect summary for covariate model

```{r}

tab_model(AUC_mel_eve, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)

```

### Compute LMM results for sparse model (relative to "dim" condition)

```{r }

summary(AUC_mel_eve_sparse)
confint(AUC_mel_eve_sparse)

```

### Random effect summary for sparse  model

```{r}

tab_model(AUC_mel_eve_sparse, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)

```


## Melatonin Onset calculations

-    Here we use data  from fitted melatonin curves resulting from the "HockeyStick" fitting method.
-    n=20 (2 participants excluded)


```{r Hockey Stick analysis, echo=FALSE}
### Fitted melatonin curves with HockeyStick----------------------------------------------------

load(file = here("LMM analysis/Datasets", "merged_mel_timing_HS.rda"))

#exclude 2 participants
merged_mel_timing_HS <- merged_mel_timing_HS %>% filter(!is.na(HS_time))


```

### Defining LMMs:

- covariate model (additionally including bright light history, pubertal stage and chronotype)

```{r }
HS_MLMO <- lmer(
formula =  HS_time ~ Block +  D1_TAT1k + mctq_sc4_chronotype + pub_stage + (1 | record_id),
                data = merged_mel_timing_HS) #  

```

- sparse model 

```{r }
HS_MLMO_sparse <- lmer(
formula =  HS_time ~ Block + (1 | record_id),
                data = merged_mel_timing_HS) #  

```

### Compute LMM results for covariate model (relative to "dim" condition)

```{r }

summary(HS_MLMO)

confint(HS_MLMO)

```

### Random effect summary for covariate model

```{r}


tab_model(HS_MLMO, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)
         

```

### Compute LMM results for sparse model (relative to "dim" condition)

```{r}

summary(HS_MLMO_sparse)
confint(HS_MLMO_sparse)

```

### Random effect summary for sparse  model

```{r}

tab_model(HS_MLMO_sparse, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)


```


## Morning melatonin AUC (as indicator for mel offset)

 - This is included in the Supplementary materials not in the main results.
 - n=20 (2 participants excluded + 1 morning AUC= NA)

### Defining LMMs

```{r }



# Define  the models with different combinations of covariates
covariates_combinations_mor <- list(
  "model0" = auc_mel_mor ~ Block + (1 | record_id),
  "model7" = auc_mel_mor ~ Block + D1_TAT1k  + mctq_sc4_chronotype + pub_stage + (1 | record_id))

```

```{r }
# Fit the models
model_results_mor <- lapply(covariates_combinations_mor, function(formula) {
  lmer(formula, data = Mel_summary)
})


```

### Compute LMM results for covariate model (relative to "dim" condition)

```{r}

# Fit a linear mixed-effects model for the summary data (no covariates)


summary(model_results_mor$model7)
confint(model_results_mor$model7)

```

### Random effect summary for covariate model

```{r }

tab_model(model_results_mor$model7, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)


```

### Compute LMM results for sparse model (relative to "dim" condition)

```{r}

summary(model_results_mor$model0)
confint(model_results_mor$model0)

```

### Random effect summary for sparse  model

```{r, echo=FALSE }

tab_model(model_results_mor$model0, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)


```


# KSS analysis

- n=4 participants excluded from KSS analysis, as they only had a third of the recorded data.
- in the KSS analysis, we use factor "time" in the LMM models (also in interaction with the condition ("Block")


```{r KSS analysis, echo=FALSE}


load(file = here("LMM analysis/Datasets", "eve_KSS.rda"))

load(file = here("LMM analysis/Datasets", "arvo_KSS.rda"))

```

## Defining LMMs:

- define 4 models: 1 covariate and  1sparse model each for analyses during the afternoon light intervention and the later evening light condition. 

- covariate models additionally include bright light history, pubertal stage and chronotype.


```{r}

KSS_eve <- lmer(
formula =  kss ~ Block * centered_time + D1_TAT1k + mctq_sc4_chronotype  + pub_stage + (1 | record_id),
                data = eve_KSS) #  

KSS_arvo <- lmer(
formula =  kss ~ Block * centered_time + D1_TAT1k + mctq_sc4_chronotype  + pub_stage + (1 | record_id),  data = arvo_KSS) #  



KSS_eve_sparse <- lmer(
formula =  kss ~ Block * centered_time + (1 | record_id),
                data = eve_KSS) #  


KSS_arvo_sparse <- lmer(
formula =  kss ~ Block * centered_time + (1 | record_id),  data = arvo_KSS) #  



```

## KSS in the later evening condition

### Compute LMM results for covariate model (relative to "dim" condition)

```{r}

summary(KSS_eve)

confint(KSS_eve)

```

### Random effect summary for covariate model

```{r}

tab_model(KSS_eve, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)


```

### Compute LMM results for sparse model (relative to "dim" condition)

```{r }


summary(KSS_eve_sparse)

confint(KSS_eve_sparse)


```

### Random effect summary for sparse  model

```{r}

tab_model(KSS_eve_sparse, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)


```

## KSS during the AEE light intervention

### Compute LMM results for covariate model (relative to "dim" condition)

```{r }

# Fit a linear mixed-effects model for the summary data (no covariates)


summary(KSS_arvo)

confint(KSS_arvo)

```

### Random effect summary for covariate model

```{r}

tab_model(KSS_arvo, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)

```

### Compute LMM results for sparse model (relative to "dim" condition)

```{r }

# Fit a linear mixed-effects model for the summary data (no covariates)


summary(KSS_arvo_sparse)

confint(KSS_arvo_sparse)


```

### Random effect summary for sparse  model

```{r}

tab_model(KSS_arvo_sparse, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)

```


# PVT analysis
-  PVT mean response speed (mean 1/RT) was the outcome of interest
- n=22 (no participants excluded)
- in the PVT analysis, we use factor "time" in the LMM models (also in interaction with the condition ("Block")


```{r, echo=FALSE}


load(file = here("LMM analysis/Datasets", "eve_PVT_stat.rda"))

load(file = here("LMM analysis/Datasets", "arvo_PVT_stat.rda"))


datastat_eve_mean_1divRT <- subset(eve_PVT_stat, variable == "mean_1divRT")

datastat_arvo_mean_1divRT <- subset(arvo_PVT_stat, variable == "mean_1divRT")


count_df <- datastat_eve_mean_1divRT %>%
  group_by(record_id,cond) %>%
  summarize(count = n())


```

## Defining LMMs:

- define 4 models: 1 covariate and  1sparse model each for analyses during the afternoon light intervention and the later evening light condition. 

- covariate models additionally include bright light history, pubertal stage and chronotype.

```{r}

PVT_mean_1divRT_eve <- lmer(
formula =  value ~ cond * centered_time + D1_TAT1k + mctq_sc4_chronotype + pub_stage + (1 | record_id),
                data = datastat_eve_mean_1divRT) #  




PVT_mean_1divRT_arvo <-  lmer(
formula =  value ~ cond * centered_time + D1_TAT1k + mctq_sc4_chronotype + pub_stage +  (1 | record_id),  data = datastat_arvo_mean_1divRT) #  



PVT_mean_1divRT_eve_sparse <- lmer(
formula =  value ~ cond * centered_time  + (1 | record_id),
                data = datastat_eve_mean_1divRT) #  




PVT_mean_1divRT_arvo_sparse  <-  lmer(
formula =  value ~ cond * centered_time + (1 | record_id),  data = datastat_arvo_mean_1divRT) #  




```

## PVT in the later evening condition

### Compute LMM results for covariate model (relative to "dim" condition)

```{r }

# Fit a linear mixed-effects model for the summary data (no covariates)

summary(PVT_mean_1divRT_eve)
confint(PVT_mean_1divRT_eve)

```

### Random effect summary for covariate model

```{r}

tab_model(PVT_mean_1divRT_eve, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)


```

### Compute LMM results for sparse model (relative to "dim" condition)

```{r }

# Fit a linear mixed-effects model for the summary data (no covariates)


summary(PVT_mean_1divRT_eve_sparse)

confint(PVT_mean_1divRT_eve_sparse)
```

### Random effect summary for sparse  model

```{r}

tab_model(PVT_mean_1divRT_eve_sparse, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)


```

## PVT during the AEE light intervention

### Compute LMM results for covariate model (relative to "dim" condition)

```{r }

summary(PVT_mean_1divRT_arvo)

confint(PVT_mean_1divRT_arvo)

```

### Random effect summary for covariate model

```{r}

tab_model(PVT_mean_1divRT_arvo, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)


```

### Compute LMM results for sparse model (relative to "dim" condition)

```{r }

# Fit a linear mixed-effects model for the summary data (no covariates)


summary(PVT_mean_1divRT_arvo_sparse)

confint(PVT_mean_1divRT_arvo_sparse)

```

### Random effect summary for sparse  model

```{r}

tab_model(PVT_mean_1divRT_arvo_sparse, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)

```


# DPG analysis

- DPG in 30 minute bins
- n=21 (1 participant excluded)
- in the DPG analysis, we use factor "time" in the LMM models (also in interaction with the condition ("Block")


```{r, echo=FALSE }


load(file = here("LMM analysis/Datasets", "eve_DPG.rda"))

load(file = here("LMM analysis/Datasets", "arvo_DPG.rda"))

```


## Defining LMMs:

- define 4 models: 1 covariate and  1sparse model each for analyses during the afternoon light intervention and the later evening light condition. 

- covariate models additionally include bright light history, pubertal stage and chronotype.

```{r }

DPG_eve <-  lmer(
formula =  DPG ~ Block * centered_time + D1_TAT1k + mctq_sc4_chronotype + pub_stage + (1|record_id),
                data = eve_DPG) # 


DPG_eve_sparse <-  lmer(
formula =  DPG ~ Block * centered_time + (1|record_id),
                data = eve_DPG) # 



DPG_arvo <-  lmer(
formula =  DPG ~ Block * centered_time + D1_TAT1k + mctq_sc4_chronotype  + pub_stage + (1|record_id),  data = arvo_DPG) #  



DPG_arvo_sparse <-  lmer(
formula =  DPG ~ Block * centered_time + (1|record_id),  data = arvo_DPG) #  




```

## DPG in the later evening condition

### Compute LMM results for covariate model (relative to "dim" condition)

```{r }

summary(DPG_eve)

confint(DPG_eve)

```

### Random effect summary for covariate model

```{r}

tab_model(DPG_eve, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)

```

### Compute LMM results for sparse model (relative to "dim" condition)

```{r }

summary(DPG_eve_sparse)

confint(DPG_eve_sparse)

```

### Random effect summary for sparse  model

```{r}

tab_model(DPG_eve_sparse, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)
```


## DPG during the AEE light intervention

### Compute LMM results for covariate model (relative to "dim" condition)

```{r }

summary(DPG_arvo)

confint(DPG_arvo)

```

### Random effect summary for covariate model

```{r}

tab_model(DPG_arvo, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)

```

### Compute LMM results for sparse model (relative to "dim" condition)

```{r }


summary(DPG_arvo_sparse)

confint(DPG_arvo_sparse)

```

### Random effect summary for sparse  model

```{r}

tab_model(DPG_arvo_sparse, show.est = FALSE, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE)

```



